---
output: html_document
---

\newcommand{\boldA}{\mathbf{A}}
\newcommand{\boldB}{\mathbf{B}}
\newcommand{\boldpi}{\boldsymbol{\pi}}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=5, echo=FALSE, message=FALSE, warnings=FALSE, eval=TRUE)
```

```{r}
library(reshape2)
library(ggplot2)
library(dplyr)

source("generate_data.R")
source("hmm.R")
source("helpers_diagnostics.R")
source("helpers_relabelling.R")
Rcpp::sourceCpp("forward_backward.cpp")
```


# HMMs: Non-Markovian data

Generate data with the hidden states $x_{1:T}$ generated as follows:

* Fix the number of breakpoints (e.g. 50)
* Draw breakpoints from the uniform distribution over $\{1, ..., n-1\}$ without replacement
* Assign class $1, ..., K=3$ to each segment independently.

### Non-Markovian Example 1

Generate observations with the emission probabilities
\[
\boldB = 
  \begin{pmatrix}
  0.5 & 0 & 0 & 0.5 \\
  0 & 0 & 0.5 & 0.5 \\
  0.6 & 0.4 & 0 & 0
  \end{pmatrix}
\]
A sequence of $x$ and $y$ of length $n=500$. 

```{r, echo=TRUE}
B = rbind(c(0.5, 0, 0, 0.5), c(0, 0, 0.5, 0.5), c(0.6, 0.4, 0, 0))

n = 500
hmm_obs = generate_nonmarkov_seq(n, B, n_breakpoints = 50)
df_true = data.frame(t = 1:n, x = hmm_obs$x)
ggplot(df_true, aes(t, x, group=1)) + 
  geom_path() + geom_point(aes(col = x)) + 
  theme_bw() + xlab("") + ggtitle("True hidden sequence")

```

#### Inference

```{r,echo=TRUE}
res = gibbs_sampling_hmm(y = hmm_obs$y, n_hidden_states = nrow(B), alpha0 = 0.1, max_iter = 1500, burnin = 500)
# relabel the latent states to obtain best match to the true B matrix
res_relabelled = match_states(res$trace_x, res$trace_A, res$trace_B, true_x = hmm_obs$x, k = nrow(B))

```


#### Visualise samples of $x_{1:T}$

Each draw from the posterior is shown as a grey line. The true states are shown as coloured points. 

```{r}
print(visualise_hidden_states(res_relabelled$trace_x, hmm_obs$x, n_display = 100))

```


#### Traceplots of the elements in $\boldB$

Panel $(i, j)$ shows the traceplot for the element $\boldB_{ij}$. The red dotted line shows the true value. 

```{r}
print(visualise_transition_mat(res_relabelled$trace_B, B))

```

#### Visualise accuracy and switch rate

```{r, fig.width=5, fig.height=5}
print(visualise_accuracy_and_switch_rate(res_relabelled$trace_x, hmm_obs$x))

```

### Non-Markovian Example 2

Generate observations with the emission probabilities
\[
\boldB = 
  \begin{pmatrix}
  0 & 0.1 & 0.35 & 0.55 \\
  0 & 0.1 & 0.45 & 0.45 \\
  0.6 & 0.3 & 0.1 & 0
  \end{pmatrix}
\]
A sequence of $x$ and $y$ of length $n=500$. 

```{r, echo=TRUE}
B = rbind(c(0, 0.1, 0.45-0.1, 0.45+0.1), c(0, 0.1, 0.45, 0.45), c(0.6, 0.3, 0.1, 0))

n = 500
hmm_obs = generate_nonmarkov_seq(n, B, n_breakpoints = 50)
df_true = data.frame(t = 1:n, x = hmm_obs$x)
ggplot(df_true, aes(t, x, group=1)) + 
  geom_path() + geom_point(aes(col = x)) + 
  theme_bw() + xlab("") + ggtitle("True hidden sequence")

```

#### Inference

```{r,echo=TRUE}
res = gibbs_sampling_hmm(y = hmm_obs$y, n_hidden_states = nrow(B), alpha0 = 0.1, max_iter = 1500, burnin = 500)
# relabel the latent states to obtain best match to the true B matrix
res_relabelled = match_states(res$trace_x, res$trace_A, res$trace_B, true_x = hmm_obs$x, k = nrow(B))

```


#### Visualise samples of $x_{1:T}$

Each draw from the posterior is shown as a grey line. The true states are shown as coloured points. 

```{r}
print(visualise_hidden_states(res_relabelled$trace_x, hmm_obs$x, n_display = 100))

```


#### Traceplots of the elements in $\boldB$

Panel $(i, j)$ shows the traceplot for the element $\boldB_{ij}$. The red dotted line shows the true value. 

```{r}
print(visualise_transition_mat(res_relabelled$trace_B, B))

```

#### Visualise accuracy and switch rate

```{r, fig.width=5, fig.height=5}
print(visualise_accuracy_and_switch_rate(res_relabelled$trace_x, hmm_obs$x))

```
