---
output: github_document
---

\newcommand{\boldA}{\mathbf{A}}
\newcommand{\boldB}{\mathbf{B}}
\newcommand{\boldpi}{\boldsymbol{\pi}}

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, echo=FALSE, message=FALSE, warnings=FALSE, eval=TRUE)
```

```{r}
library(reshape2)
library(ggplot2)
library(dplyr)
library(gridExtra)

source("generate_data.R")
source("hmm.R")
source("helpers_diagnostics.R")
source("helpers_relabelling.R")
Rcpp::sourceCpp("forward_backward.cpp")
```

### Parallel tempering for HMMs

For testing purposes, at the moment there are just parallel chains run at different temperatures, no swaps between chains happen. 

Pick temperatures $\{1, 2, 4, 8\}$ and generate the same data as for `experiment01.Rmd`. 

```{r}
# generate HMM with x in {1, 2, 3} and y in {1, 2, 3, 4}
pi = c(1/3, 1/3, 1/3)
A = rbind(c(0.9, 0, 0.1), c(0.1, 0.9, 0), c(0, 0.1, 0.9))
# B = rbind(c(0, 0.1, 0.45-0.1, 0.45+0.1), c(0, 0.1, 0.45, 0.45), c(0.6, 0.3, 0.1, 0))
B = rbind(c(0.5, 0, 0, 0.5), c(0, 0, 0.5, 0.5), c(0.6, 0.4, 0, 0))


n = 500
hmm_obs = generate_discrete_HMM(n, pi, A, B)


temperatures = c(1, 2, 4, 8)
res = parallel_tempering_hmm(y = hmm_obs$y, n_hidden_states = nrow(A), temperatures = temperatures, 
                             alpha0 = 0.1, max_iter = 1500, burnin = 500)
res_relabelled = match_states_parallel_tempering(res$trace_x, res$trace_A, res$trace_B, true_x = hmm_obs$x, k = nrow(A))

```

### Draws of $x_{1:T}$

```{r}
print(visualise_hidden_states_parallel_tempering(res_relabelled$trace_x, hmm_obs$x, temperatures, n_display = 100))

```

### Elements of the transition matrices


#### A

```{r}
visualise_transition_mat_parallel_tempering(res_relabelled$trace_A, A, temperatures)
```

#### B

```{r}
visualise_transition_mat_parallel_tempering(res_relabelled$trace_B, B, temperatures)

```


### Transition matrices without relabelling

Same as before, but do not relabel each iteration (only match the first `burnin + 1` iteration). 

```{r}
res_relabelled = match_states_parallel_tempering(res$trace_x, res$trace_A, res$trace_B, true_x = hmm_obs$x, k = nrow(A), relabel=FALSE)
```

#### A

```{r}
visualise_transition_mat_parallel_tempering(res_relabelled$trace_A, A, temperatures)
```

#### B

```{r}
visualise_transition_mat_parallel_tempering(res_relabelled$trace_B, B, temperatures)

```